The \emph{FrobeniusThresholds} package for \emph{Macaulay2}

### 2019-06-06 Fsing/FThresholdsPaper/FThresholds.tex
Updates to Section 2 of the paper

\usepackage{fancyvrb}
\usepackage{hyperref}
\hypersetup{
bookmarks,
bookmarksdepth=3,
bookmarksopen,
bookmarksnumbered,
pdfstartview=FitH,
colorlinks,backref,hyperindex,
linkcolor=Sepia,
anchorcolor=BurntOrange,
citecolor=MidnightBlue,
citecolor=OliveGreen,
filecolor=BlueViolet,
menucolor=Yellow,
urlcolor=OliveGreen
}
\usepackage{xspace}
%\usepackage{rotating}
\interfootnotelinepenalty=100000

@@

@@

We conclude this subsection by briefly reviewing some natural generalizations.  Suppose that $I$ and $J$ are ideals.  If $I$ is contained in the radical of $J$, then we set
%
\[ \nu_I^J(p^e) = \max \{ n : I^n \notin J^{[p^e]} \}. \]  
Otherwise, we set $\nu_I^J(p^e) = \infty$.

This clearly generalizes the quantity $\nu_f^{\idealm}(p^e)$ considered earlier, and we call 
\[ \ft^J(I) := \lim_{e \to \infty} \frac{ \nu_f^{\idealm}(p^e)}{p^e} \]
the \emph{$F$-threshold of $I$ with respect to $J$}.  These numbers are also rational, when finite, and we call the quantity $\ft^{\idealm}(I)$ the $F$-pure threshold of the ideal $I$.  Like $F$-pure thresholds, the $F$-thresholds of $I$ can also be defined in terms of its test ideals.

@@

There are three main sections in this paper, which describe and illustrate the main functionality implemented in the package \emph{FrobeniusThresholds}.
\begin{itemize}
\item \autoref{sec.Nu} describes the methods for approximating the $F$-pure threshold and other $F$-thresholds.  
\item \autoref{sec.IsFPT} explains the functionality for determining whether a given number is the $F$-pure threshold, or more generally, an \emph{$F$-jumping exponent}.
\item \autoref{sec.FPT} describes the features of the central function of the package, {\tt fpt}. 
\end{itemize}
This package builds heavily upon the \emph{TestIdeals} package for \emph{Macaulay2}, which provides a broad range of functionality in prime characteristic commutative algebra.

@@

Many thanks go to the organizers of the \emph{Macaulay2} workshops where much of the functionality described herein was developed, hosted by Wake Forest University in 2012, the University of California, Berkeley in 2014 and 2017, Boise State University in 2015, and the University of Utah in 2016.

Finally, we are also grateful to the University of Utah for hosting a collaborative development visit in 2018, and to the Institute of Mathematics and its Applications for its generous support for our 2019 Coding Sprint. 
The current version of the package was finalized during these events.

@@

We now consider the {\tt nu} function, which is fundamental.   As in the introduction, we work in a polynomial ring over a finite field of characteristic $p>0$.

Fix a natural number $e$.  If $f$ is polynomial contained in the radical of an ideal $J$, then the command {\tt nu(e,f,J)} outputs $\nu_f^J(p^e)$, the maximal integer exponent such that $f^n$ is not contained in $J^{[p^e]}$.  More generally, if $I$ is an ideal contained in the radical of $J$, then {\tt nu(e,I,J)} outputs $\nu_I^J(p^e)$, the maximal integer exponent $n$ such that $I^n$ is not contained in $J^{[p^e]}$.

@@

o6 = 10
\end{MyVerbatim}
}

@@

We highlight some degenerate cases:  If $f$ is zero, then the output of {\tt nu(e,f,J)} is {\tt 0}, and if $f$ is not contained in the radical of $J$, then the output is instead {\tt infinity}.  These degenerate cases extend analogously to {\tt nu(e,I,J)} when $I$ is an ideal.

@@

o8 : InfiniteNumber
\end{MyVerbatim}
}

@@

It is possible to omit the third argument from the function {\tt nu};  in this case, the function assumes that the omitted third argument is the maximal generated by the variables in the ambient polynomial ring.

@@

o12 = true
\end{MyVerbatim}
}

@@

It is well-known that if $q=p^e$ for some nonnegative integer $e$, then \[ \nu_I^J(qp) = \nu_I^J(q)\cdot p + L\]  where the error term $L$ is nonnegative, and can be explicitly bounded from above in terms of $p$ and the number of generators of $I$ and $J$.  For instance, the error term $L$ is at most $p-1$ when $I$ is principal and $J$ is arbitrary.

This implies that when searching for the maximal exponent defining {\tt nu(e,I,J)} when $e$ is positive, it is always safe to start at $p$ times the output of {\tt nu(e-1,I,J)}, and one needn't search too far past this number.  This also suggests that the most efficient way to compute {\tt nu(e,I,J)} is to compute, successively, {\tt nu(s,I,J)}, for each {\tt s = 0,\ldots,e}.  This is indeed how the computation is done in most cases.

{\color{red} Make this consistent with the introduction.  The point is that we have already described how to get the {\tt nu}-s from their limit.  Maybe we should directly cite }.  
If $M$ is the homogeneous maximal ideal of $R$ and $f$ is an element of $R$, the numbers $\nu_f^M(p^e)$ determine and are determined by the $F$-pure threshold of $f$ at the origin.  Indeed, $\nu_f^M(p^e)$ is $p^e$ times the truncation of the non-terminating base $p$ expansion of fpt($f$) at its $e^{\textrm{th}}$ spot.  This fact is used to speed up the computations for certain polynomials whose $F$-pure thresholds can be quickly computed via special algorithms, namely diagonal polynomials and binomials.  This feature can be disabled by setting the option {\tt UseSpecialAlgorithms} (default value {\tt true}) to {\tt false}.

@@

The valid values for the option {\tt ContainmentTest} are {\tt FrobeniusPower}, {\tt FrobeniusRoot}, and {\tt StandardPower}. The default value of this option depends on what is passed to {\tt nu}. Indeed, by default, {\tt ContainmentTest} is set to {\tt FrobeniusRoot} if {\tt nu} is passed a ring element $f$, and is set to {\tt StandardPower} if {\tt nu} is passed an ideal $I$.  We describe the consequences of setting {\tt ContainmentTest} to each of these values below.

First, if {\tt ContainmentTest} is set to {\tt StandardPower}, then the ideal containments checked when computing {\tt nu(e,I,J)} are verified directly.  That is, the standard power $I^n$ is first computed, and a check is then run to see if it lies in the $p^e$-th Frobenius power of $J$.

Alternately, if {\tt ContainmentTest} is set to {\tt FrobeniusRoot}, then the ideal containments are verified using Frobenius Roots.  That is, the $p^e$-th Frobenius root of $I^n$ is first computed, and a check is then run to see if it lies in $J$.  The output is unaffected, but this option often speeds up computations, specially when a polynomial or principal ideal is passed as the second argument.

@@

Finally, when {\tt ContainmentTest} is set to {\tt FrobeniusPower}, then instead of producing the invariant $\nu_I^J(p^e)$ as defined above, {\tt nu(e,I,J,ContainmentTest=>FrobeniusPower)} instead outputs the maximal integer $n$ such that the $n^{\textrm{th}}$ (generalized) Frobenius power of $I$ is not contained in the $p^e$-th Frobenius power of $J$.

@@

i1 : R = ZZ/11[x, y, z]/( x^2 - y*( z - 1 ) );

i2 : compareFPT( 5/11, z - 1 )

@@

i3 : isFPT( 1/2, z - 1 )

@@

i4 : isFPT( 1/2, z - 1, IsLocal => true )

o4 : false

@@

i50 : f = y*( ( y + 1 ) - ( x - 1 )^2 )*( x - 2 )*( x + y - 2 );

i51 : isFJumpingExponent( 3/4, f )

o51 = true

\newpage
\section{The {\tt fpt} function}
\label{sec.FPT}

@@

i2 : fpt( x^3 + y^3 + z^3 + x*y*z )

@@

i3 : fpt( x^5 + y^6 + z^7 + (x*y*z)^3 )

@@

o3 : List

@@

i4 : fpt( x^17 + y^20 + z^24 ) -- a diagonal polynomial

@@

i5 : fpt( x^2*y^6*z^10 + x^10*y^5*z^3 ) -- a binomial

@@

i7 : fpt( x^2*y^6*(x + y)^9*(x + 3*y)^10 ) -- a form in two variables

@@

o7 : QQ

@@

i10 : fpt(L, m)==o7

o10 = true

@@

i12 : fpt( f, Attempts => 0 ) -- a bad estimate

@@

i13 : fpt( f, Attempts => 0, DepthOfSearch => 3 ) -- a better estimate

@@

i14 : fpt( f, Attempts => 1, DepthOfSearch => 3 ) -- the right-hand

@@

i16 : fpt( f, Attempts => 1, DepthOfSearch => 3 )

@@

i17 : fpt( f, Attempts => 2, DepthOfSearch => 3 ) -- the left-hand endpoint

@@

i19 : fpt( f, DepthOfSearch => 3, Attempts => 2 )

@@

i20 : fpt( f, DepthOfSearch => 3, Attempts => 3 ) -- an additional check

@@

i21 : fpt( f, DepthOfSearch => 3, Attempts => 4 ) -- and one more finds

@@

o21 : QQ

@@

i23 : fpt( f, DepthOfSearch => 3 )

@@

i24 : fpt( f, DepthOfSearch => 3, UseFSignature => true )

@@

o26 : List

@@

i29 : fpt( f, Attempts => 2, DepthOfSearch => 3, UseFSignature => true ) --

      UseFSignature improves the answer a bit

@@

i30 : fpt( f, Attempts => 2, DepthOfSearch => 3, UseFSignature => true,

      FRegularityCheck => true ) -- FRegularityCheck finds the answer

@@

o30 : QQ

@@

i32 : timing numeric fpt( f, DepthOfSearch => 3, UseFSignature => true,

      FRegularityCheck => true )

@@

i33 : timing numeric fpt( f, Attempts => 5, DepthOfSearch => 3 ) -- a better

@@

i34 : timing fpt( f, DepthOfSearch => 4 ) -- the exact answer in even less
      time

@@

o34 : Time

@@

i36 : fpt( f, DepthOfSearch => 3, UseFSignature => true, Verbose => true )

@@

o36 : List

\newpage
\bibliographystyle{skalpha}
\bibliography{MainBib}

>>>>

\usepackage{fancyvrb}
\usepackage{hyperref}
\hypersetup{
bookmarks,
bookmarksdepth=3,
bookmarksopen,
bookmarksnumbered,
pdfstartview=FitH,
colorlinks,backref,hyperindex,
linkcolor=Sepia,
anchorcolor=BurntOrange,
citecolor=MidnightBlue,
citecolor=OliveGreen,
filecolor=BlueViolet,
menucolor=Yellow,
urlcolor=OliveGreen
}
\usepackage{xspace}
\interfootnotelinepenalty=100000

@@

% \begin{center}
% {\large \textbf{The \emph{FrobeniusThresholds} package for \emph{Macaulay2}}}
% \end{center}

@@

We conclude this subsection by briefly reviewing some natural generalizations.  Suppose that $I$ and $J$ are ideals.  If $I$ is contained in the radical of $J$, then we set
%
\[ \nu_I^J(p^e) = \max \{ n : I^n \notin J^{[p^e]} \}. \]  
Otherwise, we set $\nu_I^J(p^e) = \infty$.  
This clearly generalizes the quantity $\nu_f^{\idealm}(p^e)$ considered earlier, and we call 
\[ \ft^J(I) := \lim_{e \to \infty} \frac{ \nu_f^{\idealm}(p^e)}{p^e} \]
the \emph{$F$-threshold of $I$ with respect to $J$}.  These numbers are also rational, when they are finite.
If $I = (f)$ is principal, then we call $\ft^{J}(f) := \ft^J(I)$  the \emph{$F$-threshold of $f$ with respect to $J$}, 
and $\ft^{\idealm}(I)$ is the \emph{$F$-pure threshold of $I$}. 
Like $F$-pure thresholds, the $F$-thresholds of $I$ can also be defined in terms of its test ideals.

@@

There are three main sections in this paper, which describe and illustrate the main functionality implemented in the package \emph{FrobeniusThresholds}.
\begin{itemize}
\item[$\circ$] \autoref{sec.Nu} describes the methods for approximating the $F$-pure threshold and other $F$-thresholds.  
\item[$\circ$] \autoref{sec.IsFPT} explains the functionality for determining whether a given number is the $F$-pure threshold, or more generally, an \emph{$F$-jumping exponent}.
\item[$\circ$] \autoref{sec.FPT} describes the features of the central function of the package, {\tt fpt}. 
\end{itemize}
This package builds heavily upon the \emph{TestIdeals} package for \emph{Macaulay2}, which provides a broad range of functionality in prime characteristic commutative algebra.

@@

Thanks go to the organizers of the \emph{Macaulay2} workshops where much of the functionality described herein was developed, hosted by Wake Forest University in 2012, the University of California, Berkeley in 2014 and 2017, Boise State University in 2015, and the University of Utah in 2016.

Finally, we are grateful to the University of Utah for hosting a collaborative development visit in 2018, and to the Institute of Mathematics and its Applications for its generous support for our 2019 Coding Sprint. 
The current version of the package was finalized during these events.

@@

We first consider the {\tt nu} function, a fundamental component of the package \emph{FrobeniusThresholds}.   As in the introduction, in this section we work in a polynomial ring over a finite field of characteristic $p>0$.

Fix a natural number $e$.  If $f$ is a nonzero polynomial contained in the radical of an ideal $J$, then the command {\tt nu(e,f,J)} outputs the integer $\nu_f^J(p^e)$ defined in the introduction, and 
%, the maximal integer exponent such that $f^n$ is not contained in $J^{[p^e]}$.  
more generally, if a nonzero ideal $I$ is contained in the radical of $J$, then {\tt nu(e,I,J)} outputs $\nu_I^J(p^e)$. Consider the following example.  %, the maximal integer exponent $n$ such that $I^n$ is not contained in $J^{[p^e]}$.

@@

o6 = 10
\end{MyVerbatim}
}
\medspace

@@

We highlight some degenerate cases:  If $f$ is the zero polynomial, then {\tt nu(e,f,J)} outputs {\tt 0}, and if $f$ is \emph{not} contained in the radical of $J$, then the output is instead {\tt infinity}.  These degenerate cases extend analogously to the command {\tt nu(e,I,J)}.

@@

o8 : InfiniteNumber
\end{MyVerbatim}
}
\medspace

@@

If the third argument is omitted from the function {\tt nu}, the third argument is assumed to be the homogeneous maximal ideal of the polynomial ring.

@@

o12 = true
\end{MyVerbatim}
}
\medspace

@@

It is well known that if $q=p^e$ for some nonnegative integer $e$, then \[ \nu_I^J(qp) = \nu_I^J(q)\cdot p + L\]  where the error term $L$ is nonnegative, and can be explicitly bounded from above in terms of $p$ and the number of generators of $I$ and $J$.  For instance, the error term $L$ is at most $p-1$ when $I$ is principal and $J$ is arbitrary.

This implies that when searching for the maximal exponent defining {\tt nu(e,I,J)} for positive $e$, it is safe to start at $p$ times the output of {\tt nu(e-1,I,J)}, and one needn't search too far past this number.  This also suggests that the most efficient way to compute {\tt nu(e,I,J)} is to compute, successively, {\tt nu(s,I,J)}, for each $s = 0,\ldots,e$.  This is indeed how the computation is done in most cases.

{\color{red} Make this consistent with the introduction.  The point is that we have already described how to get the {\tt nu}-s from their limit.  Maybe we should directly cite }.

As pointed out in the introduction, if $\mathfrak{m}$ is the homogeneous maximal ideal and $f \in \mathfrak{m}$, then for all $e$, the values $\nu^{\mathfrak{m}}_f(p^e)$ can be recovered from the $F$-pure threshold of $f$.  
This fact is used to speed up computations for certain polynomials whose $F$-pure thresholds can be computed quickly via specialized algorithms, namely diagonal polynomials, binomials, and forms in two variables (see Section \ref{sec.FPT}).  This feature can be disabled by setting the option {\tt UseSpecialAlgorithms} (default value {\tt true}) to {\tt false}.

@@

The option {\tt ContainmentTest} for the function {\tt nu} can be set to {\tt FrobeniusPower}, {\tt FrobeniusRoot}, or {\tt StandardPower}. The default value of this option depends on what is passed to {\tt nu}. Indeed, by default, {\tt ContainmentTest} is set to {\tt FrobeniusRoot} if the second argument passed to {\tt nu} is a polynomial, and is set to {\tt StandardPower} if it is an ideal.  We describe the consequences of setting {\tt ContainmentTest} to each of these values below, using an ideal $I$ as the second argument of $\nu$, since the option works analogously when $I$ is replaced with a polynomial $f$.

First, if {\tt ContainmentTest} is set to {\tt StandardPower}, then the ideal containments checked when computing {\tt nu(e,I,J)} are verified directly.
That is, a standard power of $I$ is first computed, say $I^n$, and a check is then run to see if the resulting ideal lies in the $p^e$-th Frobenius power of $J$.
Alternately, if {\tt ContainmentTest} is set to {\tt FrobeniusRoot}, then the ideal containments are verified through the computation of Frobenius roots.  That is, the $p^e$-th Frobenius root of $I^n$ is first computed, and a check is then run to see if it lies in $J$.  The output is unaffected, but this option often speeds up computations, especially when a polynomial or principal ideal is passed as the second argument.

@@

Finally, suppose that {\tt ContainmentTest} is set to {\tt FrobeniusPower}.  Instead of producing the invariant $\nu_I^J(p^e)$, in this case the {\tt nu} function instead outputs the maximal integer $n$ such that the $n^{\textrm{th}}$ (generalized) Frobenius power of $I$ is not contained in the $p^e$-th Frobenius power of $J$ \cite{hernandez+etal.frobenius_powers}.  Note that this coincides with $\nu_I^J(p^e)$ when $I$ is replaced with a polynomial, or is a principal ideal.
After appropriately normalizing and taking the limit of these values as $e$ approaches infinity, these new values limit to certain \emph{critical exponents} (see \emph{loc.~cit.}).

@@

i1 : R = ZZ/11[x, y, z]/(x^2 - y*(z - 1));

i2 : compareFPT(5/11, z - 1)

@@

i3 : isFPT(1/2, z - 1)

@@

i4 : isFPT(1/2, z - 1, IsLocal => true)

o4 = false
\end{MyVerbatim}
}
\medspace

@@

i50 : f = y*((y + 1) - (x - 1)^2)*(x - 2)*(x + y - 2);

i51 : isFJumpingExponent(3/4, f)

o51 = true
\end{MyVerbatim}
}
\medspace

\section{The {\tt fpt} function}
\label{sec.FPT}

@@

i2 : fpt(x^3 + y^3 + z^3 + x*y*z)

@@

i3 : fpt(x^5 + y^6 + z^7 + (x*y*z)^3)

@@

o3 : List
\end{MyVerbatim}
}
\medspace

@@

i4 : fpt(x^17 + y^20 + z^24) -- a diagonal polynomial

@@

i5 : fpt(x^2*y^6*z^10 + x^10*y^5*z^3) -- a binomial

@@

i7 : fpt(x^2*y^6*(x + y)^9*(x + 3*y)^10) -- a form in two variables

@@

o7 : QQ
\end{MyVerbatim}
}
\medspace

@@

i10 : fpt(L, m) == o7

o10 = true
\end{MyVerbatim}
}
\medspace

@@

i12 : fpt(f, Attempts => 0) -- a bad estimate

@@

i13 : fpt(f, Attempts => 0, DepthOfSearch => 3) -- a better estimate

@@

i14 : fpt(f, Attempts => 1, DepthOfSearch => 3) -- the right-hand

@@

i16 : fpt(f, Attempts => 1, DepthOfSearch => 3)

@@

i17 : fpt(f, Attempts => 2, DepthOfSearch => 3) -- the left-hand endpoint

@@

i19 : fpt(f, DepthOfSearch => 3, Attempts => 2)

@@

i20 : fpt(f, DepthOfSearch => 3, Attempts => 3) -- an additional check

@@

i21 : fpt(f, DepthOfSearch => 3, Attempts => 4) -- and one more finds

@@

o21 : QQ
\end{MyVerbatim}
}
\medspace

@@

i23 : fpt(f, DepthOfSearch => 3)

@@

i24 : fpt(f, DepthOfSearch => 3, UseFSignature => true)

@@

o26 : List
\end{MyVerbatim}
}
\medspace

@@

i29 : fpt(f, Attempts => 2, DepthOfSearch => 3, UseFSignature => true)

      -- UseFSignature improves the answer a bit

@@

i30 : fpt(f, Attempts => 2, DepthOfSearch => 3, UseFSignature => true,

      FRegularityCheck => true) -- FRegularityCheck finds the answer

@@

o30 : QQ
\end{MyVerbatim}
}
\medspace

@@

i32 : timing numeric fpt(f, DepthOfSearch => 3, UseFSignature => true,

      FRegularityCheck => true)

@@

i33 : timing numeric fpt(f, Attempts => 5, DepthOfSearch => 3) -- a better

@@

i34 : timing fpt(f, DepthOfSearch => 4) -- the exact answer in even less
      time

@@

o34 : Time
\end{MyVerbatim}
}
\medspace

@@

i36 : fpt(f, DepthOfSearch => 3, UseFSignature => true, Verbose => true)

@@

o36 : List
\end{MyVerbatim}
}
\medspace

\bibliographystyle{skalpha}
\bibliography{MainBib}
